A tool set for random number generation on GPUs in R

### 2021-12-30 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
patrick edits starting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3078 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by utilizing the \pkg{clRNG} (OpenCL) library. Parallel processing by graphics processing unit (GPU) can be used to speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  %There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. 
\pkg{clrng} enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
When a RNG is called in parallel processes or successively called at several places in a program in \proglang{R}, random number generation would be more complicated because there is no guarantee that the streams (analogous to \code{.Random.seed} in \proglang{R}) do not overlap, and thus the generated sequences of random numbers is possible to have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by $Z$ steps to find its initial state, and the three states in the stream object are set to it.

>>>>

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} which leverages the \pkg{gpuR} package to generate random numbers on a Graphics Processing Unit (GPU) with the \pkg{clRNG} (OpenCL) library. Parallel processing with GPU's can speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  %There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. 
\pkg{clrng} enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

\begin{document}
\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}

@@

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
When a RNG is called in parallel processes or successively called at several places in a program in \proglang{R}, random number generation would be more complicated because there is no guarantee that the streams (analogous to \code{.Random.seed} in \proglang{R}) do not overlap, and thus the generated sequences of random numbers is possible to have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by $Z$ steps to find its initial state, and the three states in the stream object are set to it. 
When a RNG is called in parallel processes or successively called at several places in a program in \proglang{R}, random number generation becomes be more complicated because there is no guarantee that the streams (or similarly \code{.Random.seed} in \proglang{R}) do not overlap, and thus the generated sequences of random numbers is possible to have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by $Z$ steps to find its initial state, and the three states in the stream object are set to it.
